\section{Software Structure}
\label{sec:software}

Various big integer arithmetic implementations have been developed as part of
this thesis, publicly available at the GitHub repository
\url{https://github.com/tossenxD/big-int}.

The repository is structured as follows: The directories \texttt{cuda},
\texttt{futhark}, and \texttt{prototypes} contains CUDA \cpp, Futhark, and C
code, respectively. Each directory includes a \texttt{README} with detailed
explanations of the setup and a \texttt{Makefile} to replicate the results of
this thesis.

\paragraph{CUDA \cpp}
The files \texttt{ker-add.cu.h} and \texttt{ker-mul.cu.h} contains the kernels
for addition and multiplication. The arithmetics are gradually optimized over
three and five versions, respectively, kept in the files for the sake of
comparison. The kernels are called from the main file \texttt{main.cu}, which
serves to generate inputs, check for errors, test the arithmetics, and run
benchmarks. The heading of the main file defines numerous parameters specifying
integer base, kernel version, what to run, etc., which can manually be adjusted
-- along with a detailed explanation of each parameter and kernel version.

The \texttt{Makefile} offers two ways to run the main file -- with either a
normal or small amount of total work. They are invoked by \texttt{\$ make} and
\texttt{\$ make small}, respectively. The main file prints benchmark results in
the form of metrics defined in section \ref{subsec:perfmet} (, and possibly
prints the results of validation tests too, if validation is enabled).

Files to run CGBN is in subdirectory \texttt{cgbn}, and have a similar setup to
the primary files.

\paragraph{Futhark}
The files \texttt{add.fut}, \texttt{sub.fut}, \texttt{mul.fut}, and
\texttt{div.fut} contains Futhark programs with big integer addition,
subtraction, multiplication, and division. The addition and multiplication has
four and three versions, respectively, similarly to the CUDA kernels. The big
integer base type can be configured in the file \texttt{helper.fut}. While the
implementations compile for all base types, the division program is nonsensical
for bases other than \texttt{u16} due to the base restrictions explained in
section \ref{subsec:divproto}.

The subdirectories \texttt{test} and \texttt{bench} contains the Futhark test
and benchmark programs. Benchmarks are straightforward, and can be run by
\texttt{\$ make add-uX} or \texttt{\$ make mul-uX}, where \texttt{X} is the bits
of the specified base type (64 is default and 16 is not supported).  The tests
includes a C file that is tricky to call -- using the \texttt{Makefile} is
recommended. They can be run with \texttt{\$ make test-uX}, where \texttt{X} is
the bits of the base type (64 is default). The benchmarks prints runtimes and
input size, and the tests prints the count of valid testcases.

\paragraph{Sequential C Prototype}
The file \texttt{div.c} contains a sequential prototype for big integer
division, that may be of interest to run experiments on. It can be called either
with random inputs (\texttt{\$ make random}) or a fixed input (\texttt{\$ make
  fixed}), which tests against GMP. Both the number of random inputs and the
given fixed input can be adjusted in the \texttt{Makefile}. The default number
of random inputs to test with is 1000, and the \texttt{README} contains
information on how they are generated. The random tests will print the inputs,
output, and correct output of failed tests, and report the number of valid tests.

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: "../main"
%%% End:
